---
title: "The Basics 2: Vectors, Lists"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Learn how to create vectors, lists, access their elements, manipulate them, etc.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
```

## Welcome 

<div class="badge bg-primary" style="margin-top:-20px;">beginner</div>

In this tutorial, you will learn: 

  * A bit more about the types of objects in R
  * How to create lists
  * The distinction between vectors and lists
  * Access elements of vectors and lists
  * Manipulate vectors and lists

## About Vectors in R

As soon as we get into statistics or data analysis, the vector has a fundamental
role to play because it can be used to store repeated observations. I think 
many would say that the vector is actually the most fundamental data type in R.

In R, there are two types of vectors: atomic vectors and lists. Atomic 
vectors can only contain one type of data, whereas lists can contain 
elements of different types. 

Atomic vectors are what we learned about in the last tutorial: 
these are our character vectors, integer vectors, numeric vectors, and logical vectors.

So now you know if you wanted to create a numeric vector you could write `c(120, 23.7, 3, 45.3)` 
or if you wanted to create a character vector, you could write `c("Welcome", "to", "ID", "529")`.
Similar goes for logical and integer vectors: `c(TRUE, FALSE, TRUE, FALSE)` and `c(5, 2, 9)`.

The constructor for an atomic vector is the `c()` function (c stands for concatenate)
we used last tutorial. 

## Creating Lists

In contrast to using `c()` for atomic vectors, to create a list, we use the `list()` function. 

Try creating a list of some different types of elements:

```{r list1, exercise = TRUE}
# Some ideas for the name of your list: 
# patient_characteristics <- list( ... ) 
# user_feedback <- list( ... )
```

```{r list1-solution}
# One could create lists like 
patient_characteristics <- list('Very healthy', 45) 
user_feedback <- list('Almost perfect', 9/10)

print(patient_characteristics)
print(user_feedback)
```

## Named Vectors and Lists

Something that can make our lists and vectors much more informative is 
by attaching names to the entries. To create a named 
vector we use the following syntax: 

```{r named-vector1, exercise = TRUE}
patient_characteristics <- list(disposition = 'Very healthy', age = 45) 
user_feedback <- list(feedback = 'Almost perfect', rating = 9/10)

print(patient_characteristics)
print(user_feedback)
```

## Lists in Lists

One of the things that makes lists so powerful is that they can also contain
lists. This is in contrast to atomic vectors, which can only 
contain singleton elements of one of the atomic types. 

```{r list2, exercise = TRUE}
composers <- list(
  list(name = "Chopin", country = "Poland"),
  list(name = "Beethoven", country = "Germany"),
  list(name = "Mozart", country = "Austria")
)

print(composers)
```

## You can't nest atomic vectors

```{r vector_error, exercise = TRUE}
# you'll find that if you try to create a vector inside a vector, it just
# flattens
c(1, 2, c(3, 4))
```

## You can get creative with it 

For example, we could have something like this:

```{r list3, exercise = TRUE, exercise.lines=22}
composers <- list(
  chopin = list(
    name = "Chopin, Frédéric", 
    country = "Poland", 
    works = c("Nocturne Op. 9 No. 2", 
              "Prelude Op. 28 No. 15", 
              "Waltz Op. 64 No. 2")),
  beethoven = list(
    name = "Beethoven, Ludwig van", 
    country = "Germany", 
    works = c("Symphony No. 5", 
              "Symphony No. 9", 
              "Piano Sonata No. 14")),
  mozart = list(
    name = "Mozart, Wolfgang Amadeus", 
    country = "Austria", 
    works = c("Symphony No. 40", 
              "The Magic Flute", 
              "Requiem"))
)

print(composers)
```

## Access Elements of Lists and Vectors

The syntax for accessing elements of lists is slightly different to that of 
vectors. With vectors, we just use the single square brackets `[ ]` to access
elements, but with lists we can use all three of single square brackets `[ ]`, 
double square brackets `[[ ]]`, or the dollar sign `$`. 

```{r list-prep}
composers <- list(
  chopin = list(
    name = "Chopin, Frédéric", 
    country = "Poland", 
    works = c("Nocturne Op. 9 No. 2", 
              "Prelude Op. 28 No. 15", 
              "Waltz Op. 64 No. 2")),
  beethoven = list(
    name = "Beethoven, Ludwig van", 
    country = "Germany", 
    works = c("Symphony No. 5", 
              "Symphony No. 9", 
              "Piano Sonata No. 14")),
  mozart = list(
    name = "Mozart, Wolfgang Amadeus", 
    country = "Austria", 
    works = c("Symphony No. 40", 
              "The Magic Flute", 
              "Requiem"))
)

patient_characteristics <- list(disposition = 'Very healthy', age = 45) 
user_feedback <- list(feedback = 'Almost perfect', rating = 9/10)
```

```{r list4, exercise = TRUE, exercise.setup = "list-prep"}
# Access the first element of composers using the single square brackets
print(composers[1])

# Access the first element of composers using the single square brackets
print(composers[[1]])

# Access the first element of composers using the dollar sign
print(composers$chopin)
```

These more or less are all saying "access the first element," but they differ 
in their behavior. When accessing elements of a list, the main difference is
that using `[[ ]]` accesses and returns the element itself, whereas `[ ]` will
return a list containing that element. 

Using `$` to access elements is often convenient because it is supported by 
autocomplete in RStudio, and it returns lists like `[ ]` does. However, 
you can only reference named elements using the `$` syntax. 

```{r list5, exercise = TRUE, exercise.setup = "list-prep"}
# You'll find that trying composers$1 will not work, but composers[[1]] would 
composers$1 
```

## Manipulate 

Some ways we might manipulate atomic vectors and lists include: 

  * Adding elements 
  * Removing elements 
  * Accessing multiple elements 
  * Changing elements
  
```{r manipulation1, exercise = TRUE, exercise.lines = 16}
# Here's an example of each with a numeric vector
my_numbers <- c(1, 2, 3, 4, 5)

# adding an element: 
my_numbers <- c(my_numbers, 6)

# removing an element:
my_numbers <- my_numbers[-1] # remove the first element

# accessing multiple elements:
print(my_numbers[c(1, 3, 5)]) # access the first, third, and fifth elements

# changing an element:
my_numbers[1] <- 10 # change the first element to 10

print(my_numbers)
```

```{r manipulation2, exercise = TRUE, exercise.lines = 21}
# Here's an example of each with a list
my_list <- list(
  name = "Chopin, Frédéric", 
  country = "France", # whoops
  works = c("Nocturne Op. 9 No. 2", 
            "Prelude Op. 28 No. 15", 
            "Waltz Op. 64 No. 2"))

# adding an element:
my_list <- c(my_list, age = 39)

# removing an element:
my_list <- my_list[-length(my_list)] # remove the last 

# accessing multiple elements:
print(my_list[c('name', 'country')]) # access the first and third elements

# changing an element:
my_list$country <- "Poland" # change the second element to France

print(my_list)
```

## Summary 

We learned some important points in this tutorial: 

  * That atomic vectors only contain one type of data, while lists are more flexible 
  * How to modify atomic vectors and lists 
  * How to access elements of lists

<div style="width:100%;height:0;padding-bottom:53%;position:relative;"><iframe src="https://giphy.com/embed/3TZgJXiwbdbLG" width="100%" height="100%" style="position:absolute" frameBorder="0" class="giphy-embed" allowFullScreen></iframe></div><p><a href="https://giphy.com/gifs/ghibli-my-neighbor-totoro-3TZgJXiwbdbLG">via GIPHY</a></p>

References 

  - R Basics: Quick & Easy <http://www.sthda.com/english/wiki/r-basics-quick-and-easy> 
  - R for Beginners, Emmanuel Paradis <https://cran.r-project.org/doc/contrib/Paradis-rdebuts_en.pdf>
  - R for Data Science, Hadley Wickham & Garrett Grolemund <https://r4ds.had.co.nz/>
