---
title: "The Basics 1: Intro to Objects and Vectors"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Learn how to create vectors, lists, access their elements, manipulate them, etc.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
```

## Welcome 

<div class="badge bg-primary" style="margin-top:-20px;">beginner</div>

In this tutorial, you will learn how to: 

  * Learn about the types of objects in R
  * Create vectors
  * Access elements of vectors
  
### Types of Objects in R

R has four basic or **"atomic"** classes of objects:

  - character
  - numeric
  - integer
  - logical

For the most part, everything more complicated than these objects is built up from these basic objects. So let's start with how to initialize these types of objects. 

Do you want to give it a try first after what you've seen from class? 

```{r objects1, exercise = TRUE}
# Create a character object called my_character with the value "Hello World!"


# Create an integer object called my_integer with the value 529


# Create a numeric object called kgs_per_pound and set it to 0.454


# Create a logical object called is_it_january and set it to TRUE

```

```{r objects1-hint}
# To declare variables, we use a syntax that looks like this:
my_character <- __________ 

# The special things you need to pay attention to for filling in the blank are that:
#   * for a character object you want to either wrap it in single or double 
#     quotes, like 'Hello World!' or "Hello World!", 
#   * and for a logical object you want to use either TRUE or FALSE in all-caps.

# The next hint is the solution — so give it a solid attempt first! 
```

```{r objects1-solution}
my_character <- "Hello World!"
my_integer <- 529
kgs_per_pound <- 0.454
is_it_january <- TRUE
```

You may notice that R doesn't provide any feedback when you create an object. 
That's OK! The feedback we're really getting here is that no error was thrown.

### A Note on Naming

<div class="alert alert-dismissible alert-info">
  <p class="mb-0">
  You may be curious what are valid names for an object in R. 
  
  A name for an object must start with a character and can contain letters,
  numbers, periods and underscores. Typically one should prefer to use
  underscores over periods because periods are used with special meaning in R's
  object oriented systems.
  </p>
</div>


### Checking Objects' Values

You'll note there were actually a few conceptual steps in creating a variable:
1) we had to give it a name, 2) we had to use an assignment operator (the `<-`
arrow), and 3) we had to specify what the assigned value would be ("Hello
World!").

Now we might want to check what the values of our objects are. 
You can do that by simply entering the name of the object into 
the R console and running it, and then R will show you the value 
of the object. 

```{r objects-prep}
my_character <- "Hello World!"
my_integer <- 529
kgs_per_pound <- 0.454
is_it_january <- TRUE
```

```{r objects2, exercise = TRUE, exercise.setup = "objects-prep"}
# try it out 
```

```{r objects2-hint}
# try just putting in my_character and clicking run 
```

If you were trying to check the value of multiple objects, you may have 
noticed that R only returned the value of the last object you entered. 
Try comparing the two output to each other:

```{r objects3, exercise = TRUE, exercise.setup = "objects-prep"}
# First, just try uncommenting these four lines and running the code:
# my_character
# my_integer
# kgs_per_pound
# is_it_january

# Second, try commenting out the above and uncommenting the next four lines and
# running the code again:
# print(my_character)
# print(my_integer)
# print(kgs_per_pound)
# print(is_it_january)
```

```{r objects3-hint}
# If you're not sure what I mean by "uncomment" or "comment out", here's what 
# I mean:

# The hashtag character (#) is used to create comments in R code — 
# That means that the R interpreter will ignore anything on a line
# after a # character appears. 

# To uncomment a line, just delete the # character at the beginning of the line.
# To comment out a line, just add a # character at the beginning of the line. 
```

### Checking the Types of Objects

Sometimes you may want to check what the type of an object in 
R is. You can use the `typeof()` or `class()` function on an 
object to check its type. 

```{r types, exercise = TRUE, exercise.setup = "objects-prep"}
```

```{r types-solution, exercise = TRUE}
typeof(my_character)
typeof(my_integer)
typeof(kgs_per_pound)
typeof(is_it_january)
```

### Just so you know

<div class="alert alert-dismissible alert-info">
  <p class="mb-0">
  R also has two other important classes, which are functions and NULL. Null values 
like `NULL`, `NA`, `NaN`, and `Inf` are also important to know about, but we want to make sure to cover the basics first. 
  </p>
</div>

### Vectors

Vectors are a specific type of object in R that can hold multiple values of the 
**same type**. So for example, a vector can have all integer values, or all 
logical values, but it can't mix two different types.

Vectors can be created using the `c()` function, which stands for "combine".

Let's give it a go:

```{r vectors1, exercise = TRUE}
# Create a vector called my_vector with the values 1, 2, 3, 4, 5

# Create a vector called my_vector2 with the values "a", "b", "c"
```

```{r vectors1-hint}
# To create a vector, we use a syntax that looks like this:
# my_vector <- c(__, __, __, __) 

# The next hint is the solution, so give it a go first!
```

```{r vectors1-solution}
my_vector <- c(1, 2, 3, 4, 5)
my_vector2 <- c("a", "b", "c")
```

### Accessing Elements of Vectors

You can access elements of vectors by using square brackets `[]` and specifying
the index of the element you want to access. 

```{r vectors-prep}
my_vector <- c(1, 2, 3, 4, 5)
my_vector2 <- c("a", "b", "c")
```

```{r vectors2, exercise = TRUE, exercise.setup = "vectors-prep"}
# Access the first element of my_vector

```

```{r vectors2-hint}
# To access the first element of a vector, we use a syntax that looks like this:
# my_vector[_]
```

```{r vectors2-solution}
my_vector[1]
```

### Length of a Vector

The length of a vector is the number of elements it contains. You can check the
length of a vector using the `length()` function.

```{r vectors3, exercise = TRUE, exercise.setup = "vectors-prep"}
# Check the length of my_vector and my_vector2

```

```{r vectors3-solution}
length(my_vector)
length(my_vector2)
```

### Last element of a vector

<br><br>

Using what you just learned, can you think of how to access the last element of
a vector?

```{r vectors4, exercise = TRUE, exercise.setup = "vectors-prep"}
# Access the last element of my_vector
```

```{r vectors4-hint}
# We can use my_vector[n] to access the nth element of my_vector.
# We can use length(my_vector) to get the length of my_vector.
# Check the next hint for the solution
```

```{r vectors4-solution}
my_vector[length(my_vector)]
```

### Accessing Any Element

You can also use a variable when referencing the index
of the vector element you want — let's try that out:

```{r vectors5, exercise = TRUE, exercise.setup = "vectors-prep"}
# Create a variable called n and set it to 3 
# Then access the nth element of my_vector2
```

```{r vectors5-solution}
n <- 3
my_vector2[n]
```

### Congrats! 

You finished this tutorial! 

<iframe src="https://giphy.com/embed/3GBG7xejOXhH7MyQkn" width="480" height="384" frameBorder="0" class="giphy-embed" allowFullScreen></iframe><p><a href="https://giphy.com/gifs/molangofficialpage-kawaii-molang-piupiu-3GBG7xejOXhH7MyQkn">via GIPHY</a></p>

References 

  - R Basics: Quick & Easy <http://www.sthda.com/english/wiki/r-basics-quick-and-easy> 
  - R for Beginners, Emmanuel Paradis <https://cran.r-project.org/doc/contrib/Paradis-rdebuts_en.pdf>
  - R for Data Science, Hadley Wickham & Garrett Grolemund <https://r4ds.had.co.nz/>

  
