---
title: "Data Frames and Matrices"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Learn how to create data frames, matrices, and perform some basic operations with them
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
```

## Welcome

<div class="badge bg-primary" style="margin-top:-20px;">beginner</div>

Another basic data structure that you will encounter all the time 
in R is the data frame. Data frames are the most common way that data is stored in R.
Data frames are, under the hood, just a list of columns (vectors) that are all the same length. 

### Creating Data Frames  

There are many ways to create data frames in R. Let's start with the 
most basic, which is to use the `data.frame()` function built into R.

(Sometimes I'll say that something is built into "base R" which 
means the same thing as built into R). 

```{r dataframes1, exercise = TRUE}
# Create a data frame called df with the following columns:
# columns: "name", "age", "height", "weight"
df <- data.frame(
  name = c(______, _____, ______),
  age = c(__, __, __),
  height = c(____, ____, ____),
  weight = c(____, ____, ____)
)
```

```{r dataframes1-solution}
# Create a data frame called df with the following columns:
# columns: "name", "age", "height", "weight"
df <- data.frame(
  name = c("Pritha", "Nikhil", "Akiyo"),
  age = c(25, 30, 35),
  height = c(1.8, 1.6, 1.7),
  weight = c(60, 80, 70)
)
```

### Accessing Data Frames

There are many ways to access data frames in R. Let's start with the
most basic, which is to use the `$` operator. 

```{r dataframes-prep}
df <- data.frame(
  name = c("Pritha", "Nikhil", "Akiyo"),
  age = c(25, 30, 35),
  height = c(1.8, 1.6, 1.7),
  weight = c(60, 80, 70)
)
```

```{r dataframes2, exercise = TRUE, exercise.setup = "dataframes-prep"}
# Access the "name" column of df using the $ operator
df$____
```

```{r dataframes2-solution}
df$name
```

### 

There are other parts of a data frame we might want to access, like we 
might want to check the names of the columns, or the 
number of rows or columns.

```{r dataframes3, exercise = TRUE, exercise.setup = "dataframes-prep"}
# Check the names of the columns of df
____(df)

# Check the number of rows and columns of df
____(df)
```

```{r dataframes3-solution}
# Check the names of the columns of df
names(df)

# Check the number of rows and columns of df
dim(df) # returns a vector equivalent to c(nrow(df), ncol(df))
nrow(df)
ncol(df)
```

### 

We can also access specific rows and columns of a data frame using
the `[` operator. 

```{r dataframes4, exercise = TRUE, exercise.setup = "dataframes-prep"}
# Access the first row of df
df[____, ]

# Access the first column of df
df[, ____]
```

```{r dataframes4-solution}
# Access the first row of df
df[1, ]

# Access the first column of df
df[, 1]
```

###

We can also access multiple rows and columns of a data frame using the `[` operator. 

```{r dataframes5, exercise = TRUE, exercise.setup = "dataframes-prep"}
# Access the first and third rows of df
df[____, ]
```

```{r dataframes5-solution}
# Access the first and third rows of df
df[c(1, 3), ]
```

### Matrices

Matrices are another common data structure in R. Matrices are like data frames,
but they can only contain one type of data (e.g. all numbers, or all strings).

### Creating Matrices

There are many ways to create matrices in R. Let's start with the
most basic, which is to use the `matrix()` function built into R.

```{r matrices1, exercise = TRUE}
# Create a matrix called mat with the following values:
# 1, 2, 3
# 4, 5, 6
# 7, 8, 9

mat <- matrix(
  c(____, ____, ____,
    ____, ____, ____,
    ____, ____, ____),
  nrow = ____, ncol = ____, byrow = ____
)
```


```{r matrices1-solution}
# Create a matrix called mat with the following values:
# 1, 2, 3
# 4, 5, 6
# 7, 8, 9

mat <- matrix(
  c(1, 2, 3,
    4, 5, 6,
    7, 8, 9),
  nrow = 3, ncol = 3, byrow = TRUE
)
```

```{r matrix-prep}
df <- data.frame(
  name = c("Pritha", "Nikhil", "Akiyo"),
  age = c(25, 30, 35),
  height = c(1.8, 1.6, 1.7),
  weight = c(60, 80, 70)
)

mat <- matrix(
  c(1, 2, 3,
    4, 5, 6,
    7, 8, 9),
  nrow = 3, ncol = 3, byrow = TRUE
)

mat2 <- matrix(
  c(1, 2, 3,
    4, 5, 6,
    7, 8, 9),
  nrow = 3, ncol = 3, byrow = TRUE, 
  dimnames = list(c("row1", "row2", "row3"), c("col1", "col2", "col3"))
)
```

### Accessing Matrices

Similar to data frames, there are many ways to access matrices in R. Let's start with the
most basic, which is to use the `[` operator. 

```{r matrices2, exercise = TRUE, exercise.setup = "matrix-prep"}
# Access the last row of mat
mat[____, ]

# Access the last column of mat
mat[, ____]
```

```{r matrices2-solution}
# Access the last row of mat
mat[nrow(mat), ] # or mat[3, ]

# Access the last column of mat
mat[, ncol(mat)] # or mat[, 3]
```

### Names for Matrices

Matrices can also have names for their rows and columns. 

```{r matrices3, exercise = TRUE, exercise.setup = "matrix-prep"}
# Create a matrix called mat2 with the following values:
# 1, 2, 3
# 4, 5, 6
# 7, 8, 9
# and with the following row and column names:
# "row1", "row2", "row3"
# "col1", "col2", "col3"

mat2 <- matrix(
  c(____, ____, ____,
    ____, ____, ____,
    ____, ____, ____),
  nrow = ____, ncol = ____, byrow = ____, 
  dimnames = list(____, ____)
)
```

```{r matrices3-solution}
# Create a matrix called mat2 with the following values:
# 1, 2, 3
# 4, 5, 6
# 7, 8, 9
# and with the following row and column names:
# "row1", "row2", "row3"
# "col1", "col2", "col3"

mat2 <- matrix(
  c(1, 2, 3,
    4, 5, 6,
    7, 8, 9),
  nrow = 3, ncol = 3, byrow = TRUE, 
  dimnames = list(c("row1", "row2", "row3"), c("col1", "col2", "col3"))
)
```

### Accessing Individual Elements

We can access individual elements of either a matrix or a data frame using the `[` operator. 

```{r matrices4, exercise = TRUE, , exercise.setup ='matrix-prep'}
# Access the element in the first row and second column of mat2
mat2[____, ____]
df[____, ____]
```

```{r matrices4-solution}
# Access the element in the first row and second column of mat2
mat2[1, 2]
df[1, 2]
```

### Operations on Matrices and Data Frames

Some of the most common operations we might want to do on matrices are
`colMeans()`, `colSums()`, `rowMeans()` and `rowSums()`.

```{r matrices5, exercise = TRUE, exercise.setup = 'matrix-prep'}
# Calculate the column means of mat2
colMeans(mat2)

# Calculate the row sums of mat2
rowSums(mat2)
```

```{r matrices5-solution}
# Calculate the column means of mat2
colMeans(mat2)

# Calculate the row sums of mat2
rowSums(mat2)
```

### Matrix Arithmetic 

We're not going to be doing a lot of matrix arithmetic in this class, but 
it's useful to have seen that matrix multiplication uses a 
special operator `%*%` in R. If you use `*` without the `%`s, you'll get
element-wise multiplication. 

```{r matrices6, exercise = TRUE, exercise.setup = "matrix-prep"}
## Remember the contents of mat2? 
mat2

## Element-wise multiplication
mat2 * mat2

## Matrix multiplication
mat2 %*% mat2
```

### Summary 

What did we learn? 

  - Data frames are the most common way that data is stored in R.
  - Data frames are, under the hood, just a list of columns (vectors) that are all the same length.
  - Matrices are like data frames, but they can only contain one type of data (e.g. all numbers, or all strings).
  - Some of the most common operations we might want to do on matrices are
`colMeans()`, `colSums()`, `rowMeans()` and `rowSums()`.
  - Matrix multiplication uses a special operator `%*%` in R. If you use `*` without the `%`s, you'll get
element-wise multiplication.

<iframe src="https://giphy.com/embed/ayBZf3xVtT74Q" width="343" height="480" frameBorder="0" class="giphy-embed" allowFullScreen></iframe><p><a href="https://giphy.com/gifs/studio-ghibli-spirited-away-ayBZf3xVtT74Q">via GIPHY</a></p>
