---
title: "Tutorial for Conditionals, Logic, and Programming Flow"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Learn how to use conditionals, logic, and programming flow in R!
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
```

### Welcome

In this tutorial, you will learn how to:

  * Use `if`, `else`, `for` and `while`
  * Use `&` and `|`
  * Know when to use `in` vs `%in%`
  * Get comfortable using `==`, `!=`, and `!`

### `if` 

`if` let's you evaluate code dependent on some condition specified in R. 

For example, the following code will print "hello world!" on the console because the condition given is `TRUE`:

```{r example-if, echo=TRUE, eval=FALSE}
if (TRUE) { 
  print("hello world!") 
}
```

How could you modify the following code to print "Battlestar Galactica" to the console?

```{r if1, exercise=TRUE, exercise.eval=FALSE}
if (_____) {
  print("Battlestar Galactica")
}
```

```{r if1-solution}
if (TRUE) {
  print("Battlestar Galactica")
}
```

### Using Meaningful Conditions

The conditions given can get more complicated.  For example, what do you think the following code might do?

```{r if2, eval = FALSE, echo = TRUE}
x <- 1
if (x > 0) {
  print("Pokemon Go!")
}
```

It prints "Pokemon Go!"

### A Few Examples

Here are some examples that are fairly common types of places `if` might be used:

```{r, eval = FALSE, echo = TRUE}
# if certain conditions require warning or distinct behavior, you might write
# something like:

# note that the process for generating x could be any data generating process,
# but here we get a random integer in 1 to 10

x <- sample(n = 10, size = 1) 

if (x > 5) {
  warning("x is greater than 5")
}
```


```{r, eval = FALSE, echo = TRUE}
# sometimes you'll encounter missing data that may need to be especially handled: 

# again, imagine this code for generating x could be a more complicated data
# generating process, like data someone gave you.

x <- c(0, 1, NA)[sample(n = 3, size = 1)] 
if (is.na(x)) {
  warning("x is missing data!")
}
# or: 
if (is.na(x)) {
  x <- 0.5 # impute x with a specific value; not often advised, but sometimes warranted
}
```

```{r, eval = FALSE, echo = TRUE}
# a more complicated example might be testing if a file exists and then reading
# it in if it actually exists:

path_to_file <- "/path/to/filename.txt"
if (file.exists(path_to_file)) {
  file_contents <- readLines(path_to_file)
}
```

### Multiple Branches 

A programmer can use `if` and `else` together to evaluate one of two programming
expressions dependent on the condition given.

For example, what do you think the following might do? 

```{r logicals, echo = FALSE}
question("What will the following code return? 
  ```r
  x <- 123
  if (is.character(x)) {
   print('The Simpsons')
  } else {
   print('Sesame Street')
  }
  ```
",
         answer("The Simpsons", message = "x isn't a character object, so this answer is incorrect."),
         answer("Sesame Street", correct = TRUE, message = "x is not a character object, so the code in the `else` block is what gets evaluated, resulting in 'Sesame Street' being printed to the console"),
         answer("Blues Clues", message = "This didn't even appear in any of the branches in the code, so there's no way this could have been the output."),
         allow_retry = TRUE
)
```

You can even use multiple `if` `else` statements chained together. 

What do you think the following might do? 

```{r elseif, echo=FALSE}
question("What will the following code return? 
  ```r
  x <- 123
  if (is.character(x)) {
   print('The Simpsons')
  } else if (is.numeric(x)) {
   print('Sesame Street')
  } else {
    print('Doctor Who')
  }
  ```
",
         answer("The Simpsons", message = "x isn't a character object, so this answer is incorrect."),
         answer("Sesame Street", correct = TRUE, message = "x is not a character object, so the code in the `else` block is what gets evaluated, resulting in 'Sesame Street' being printed to the console"),
         answer("Doctor Who", message = "x is an object of numeric type, so the branch 'Doctor Who' appears in won't be the branch of the code that's evaluated."),
         answer("Blues Clues", message = "This didn't even appear in any of the branches in the code, so there's no way this could have been the output."),
         allow_retry = TRUE
)
```

### `for` 

When we want to solve repetitive tasks (something computers & machines are great at helping with), we often turn to 
using loops in programming languages.  Loops define behavior that is repeated some programmatically determined number of times. 
The two types of loops we'll be using are `for` and `while` loops, and we'll be starting with `for` loops here. 

Let's have an example to talk about: 

```{r}
for (i in 1:5) {
  print(i)
}
```

The basic syntax will always be the same for a `for` loop — they always have these 
three components: 

  * they start with `for`
  * then there's a parenthetical with an `in` statement like (__ in ___)`
  * and they have a block of code that is repeated, i.e., what's inside the `{}`.

What the `for` loop does is it assigns the *iterator* (here named `i`) to each level
of the sequence given (here `1:5`, the values 1,2,3,4,5) and then evaluates the 
code given in the block.  So if we were to re-write the above without using `for`
it would look like this:

```{r, eval=FALSE}
i <- 1
print(i)
i <- 2
print(i)
i <- 3
print(i)
i <- 4
print(i)
i <- 5
print(i)
```

A bit more verbose right? 

`for` loops can be used with sequences besides integer sequences — as long as it
can be indexed using integers, it can be used as a sequence for a `for` loop. 

```{r for, echo=FALSE}
question("What will the following do? 
  ```r
  fruits <- c('apple', 'banana', 'persimmon', 'boysenberry')
  for (fruit in fruits) {
    print(fruit)
  }
  ```
",
         answer("Print 'apple'", message = "It will print 'apple', but the code will do more than that 🙂"),
         answer("Print 'fruit'", message = "Not quite; fruit is just a variable, so the value it represents will be what actually gets printed;  moreover, it's inside a for loop, so the fruit variable will be printed multiple times."),
         answer("Print 'apple' then 'banana' then 'persimmon' then 'boysenberry'", correct = TRUE, message = "That's right! The fruit variable will get assigned 'apple', then be printed, and so-on and so-forth for each of the values in the fruits vector."),
         answer("Print 'fruits'", message = "Not quite; 'fruits' is the name of the vector that is the sequence the `for` loop will iterate over, not what will get printed."),
         allow_retry = TRUE
)
```

### `while`

The second type of loops we'll introduce you to are `while` loops.  `while` loops are particularly useful 
when it's uncertain when some action needs to be done until a certain condition is reached and it's
difficult to determine in advance how many times that action will need to be done. 

For example, imagine you are picking apples and take out the seeds from each
apple.  You'd like to collect at least 100 apple seeds, but the number of seeds
in each apple varies (typically between 5-13 seeds per apple).

We can simulate this process in a `while` loop. 

```{r}
seeds_collected <- 0 # to start with, we haven't collected any apple seeds 

while (seeds_collected < 100) {
  
  # grab a new apple and collect its seeds
  new_seeds <- c(5:13)[sample.int(n = 9, size = 1)] # a random number between 5 to 13
  
  seeds_collected <- seeds_collected + new_seeds # now we have what we've already collected + the new seeds
}

print(seeds_collected) # at the end, we'll have 100 or more seeds!
```

The basic structure of a `while` loop are the following three components: 

  * `while` at the beginning
  * a condition inside the `()` parentheses
  * and a block of code to evaluate repeatedly until the condition given becomes `TRUE`. 

Ok, maybe collecting apple seeds doesn't sound like something that your job will
depend on, but how about a modeling problem where you need to be confident that
your model parameters have done a good job fitting the data? 

Here's how we might write that out **in pseudocode** using a `while` loop: 

```{r, eval=FALSE}
parameters <- ... # our model has some parameters we're optimizing

while (model_fit > acceptable_threshold) {
  
  # find some new acceptable, hopefully better parameters for the model
  parameters <- find_new_candidate_parameters()
  
  # evaluate how good the model performs using the new parameters
  model_fit <- measure_model_performance(parameters)
}
```

```{r while, echo=FALSE}
question("What will the following do? 
  ```r
  x <- 0
  while (x < 100) {
    x <- x + sample.int(n = 10, size = 1)
    print(x)
  }
  ```
",
         answer("Print 0", message = "Nope. `x` will have already had a value of at least 1 added to it before the first time its printed.  Moreover, it will get printed multiple times in this `while` loop."),
         answer("Print 100", message = "Nope. 100 is just the threshold where once x is 100 or greater, the code in the `{}` brackets won't be repeated anymore."),
         answer("Print 100 numbers between 0 and 100", message = "Nope; there's no guarantee that exactly 100 numbers will be printed. Because `x` is being incremented by a *random* number between 1 and 10 each time, we can't be certain how many numbers will be printed out."),
         answer("Print out some indeterminate number of numbers, at least 10, between 1 up to 109", correct = TRUE, message = "Great job figuring it out!"),
         allow_retry = TRUE
)

```


### `&` and `|`

`&` and `|` are logical operators that represent the logical "AND" and "OR" operations. 

In R and in formal logic settings, we say that `x & y` is `TRUE` if both `x` and `y` 
are true.  Similarly, we say `x | y` (read: `x` or `y`) is `TRUE` if one or
both of `x` and `y` is `TRUE`. 

Here are some examples: 

```{r, eval=F}
TRUE & TRUE # is true
TRUE & FALSE # is false 
FALSE & TRUE # is false 
FALSE & FALSE # is false 
TRUE | TRUE # is true
TRUE | FALSE # is true 
FALSE | TRUE # is true 
FALSE | FALSE # is false
```

It does sometimes get more complicated than this.  Here's a trickier example: 

```{r nested_logic, echo=FALSE}
question("What will `TRUE & (FALSE | TRUE)` evaluate to?",
         answer("`FALSE`", message = "Nope. The key insight here is that the inner expression `(FALSE | TRUE)` (read: FALSE or TRUE) will evaluate to `TRUE`, and then the outer expression evaluates as `TRUE & TRUE`, which is `TRUE`."),
         answer("`TRUE`", correct = TRUE, message = "That's correct!  It's a bit tricky, but you got it. The key insight here is that the inner expression `(FALSE | TRUE)` (read: FALSE or TRUE) will evaluate to `TRUE`, and then the outer expression evaluates as `TRUE & TRUE`, which is `TRUE`."),
         allow_retry = TRUE
)
```


### in versus %in%

You've already seen `in`, which appeared in how iteration along a sequence is specified 
in a `for` loop.  R has another, similar operator, `%in%` that **tests** for if
a given variable (typically a single value) appears inside another variable
(typically a vector or list).

That is to say, `x %in% y` tests if `x` appears in `y`.  Here's an example: 

```{r}
'apple' %in% c('apple', 'banana', 'pear', 'starfruit')
```

`%in%` will return either `TRUE` or `FALSE` depending on whether or not its left-hand-side
is in its right-hand-side (i.e., `x %in% y` will be `TRUE` if `x` is in `y`).

So why are there two?  When do you use one and not the other?  Thankfully the answer
is simple: you only need to use `in` inside a `for` loop, and `%in%` is what you use
to test if a variable is contained in another variable.

### `==`, `!=` and `!` 

First, congratulate yourself on making it this far!  This has been a lot, and
you're doing great!

Thankfully, we're wrapping up.

`==` tests for equality between two variables, and similarly `!=` tests for 
inequality. 

Some examples: 

```{r, eval=FALSE}
'apple' == 'apple' # TRUE
1 == 1 # TRUE

'apple' != 'banana' # TRUE, because 'apple' is _not_ equal to 'banana'
1 != 2 # TRUE, because 1 is _not_ equal to 2
```

**A word to the wise:** it's very easy to get `=` and `==` mixed up (especially just as a typo),
so be especially careful to make sure you don't make typos with these since
their meaning is different. `=` typically works just like the assignment `<-`
operator, so that will overwrite your variables, while `==` tests for equality
(and returns `TRUE` or `FALSE`). Since they do different things, it can be quite
an annoying bug when you accidentally make a typo swapping `==` for `=` or vice
versa.

So what about the `!` operator?  Just read it like "not". 

```{r not, echo=FALSE}
question("What will `! TRUE` evaluate to?",
         answer("`FALSE`", correct = TRUE, message = "Yes, exactly! 'Not `TRUE`' would be `FALSE` indeed." ),
         answer("`TRUE`", message = "Ah, not exactly. You should read the statement to yourself as '**NOT** `TRUE`'"),
         allow_retry = TRUE
)
```

